---
title: "Exercise 03: Explore and Wrangle Data"
format: html
execute:
  echo: true
  warning: false
  message: false
---

## Setup

tidyverse library for data import and wrangling.

```{r}
library(tidyverse)
```

## Loading the dataset and inspecting variables

From the provided URL we can first load the data set into the object called "d".

```{r}
f<- "https://raw.githubusercontent.com/difiore/ada-datasets/main/data-wrangling.csv"
d<- read_csv(f, col_names = TRUE)

```

We can take a quick look at the data we loaded and variables in it.

```{r}
head(d)
dim(d)
str(d)
glimpse(d)
types <- identify_variable_types(d)


#Listing variable names

names(d)
```

## Step 1

#### Creating new variables

#### **1) BSD** (body size dimorphism) - Ratio of average male to female body mass.

#### **2) sex_ratio** - Ratio of the number of adult females to adult males in a typical group.

#### **3) DI** (for “defensibility index”) - Ratio of day range length to the diameter of the home range.

```{r}
#BSD (body size dimorphism) - Ratio of average male to female body mass

d$BSD<-d$Body_mass_male_mean/d$Body_mass_female_mean
d$BSD

#sex_ratio - Ratio of the number of adult females to adult males in a typical group

d$sex_ratio<-d$AdultFemale/d$AdultMales
d$sex_ratio


```

We need a calculation here, to get the diameter of the home range to be derived from HomeRange_km2 in order to get the new variable "DI-defensibility index"

```{r}
d$HomeRange_diameter <- 2 * sqrt(d$HomeRange_km2 / pi)
d$HomeRange_diameter

#DI (for “defensibility index”) - Ratio of day range length to the diameter of the home range

d$DI <- d$DayLength_km/d$HomeRange_diameter
d$DI
```

Since a division is included we must make sure that the rows where the relavant denominator is zero the new variable is INF. To check this,

```{r}
which(is.infinite(d$BSD)) # Checking whether denominator zeros have given any INF
which(is.infinite(d$sex_ratio))
which(is.infinite(d$DI))
```

There is no any instances as INF.

#### 4) Relationship between day range length and time spent moving

First trying out **ggplot2** to visualize this relationship.

The reason for choosing **ggplot2** for this plot is we can use faceting by family to generate one plot per family. So our comparison will be easy.Using a scatterplot because both are continuous variables,

Explaining the layers and inside parameters used : aesthetics layer - Mapping given variables to x and y axis, to customize the plot..geometry layer - 0.5 opacity points, color=black, size=4 and a smooth line to show where all points are lying generally. Using labs() to set the x- and y-axis labels.

```{r}
library(ggplot2)

# Overall plot (all species)

ggplot(d, aes(x = Move, y = DayLength_km)) +
  geom_point(alpha=0.5,color="black",size=4, na.rm = TRUE) + geom_smooth(method = lm)+labs(x="Tiem spent moving", y="Day range length")
```

Should we use log values to plot? How to decide??

#### By family comparison

1\) Approach 1: First using the color aesthetic to show different families.

```{r}
library(ggplot2)

# plot all species in one plotwith dfferent color for each

ggplot(d, aes(x = log(Move), y = log(DayLength_km), color=Family)) +
  geom_point(alpha=0.5, na.rm = TRUE)+labs(x="Tiem spent moving", y="Day range length")
```

2\) Approach 2- Now, we are using facet_wrap() layer in ggplot2 to visualize families in a separate plot

```{r}
ggplot(d, aes(x = Move, y = DayLength_km,color = factor(Family))) +
  geom_point(na.rm = TRUE)  +
  facet_wrap(~ Family)+labs(
    x = "Time spent moving",
    y = "Day range length (km)"
  )+labs(x="Tiem spent moving", y="Day range length")
```

#### 5) relationship between day range length and group size 

```{r}
# Overall plot (all species)
ggplot(d, aes(x = MeanGroupSize, y = DayLength_km, Family= disp)) +
  geom_point(na.rm = TRUE) +
  geom_smooth(method = "lm", se = FALSE, na.rm = TRUE)
```

```{r}
ggplot(d, aes(x = MeanGroupSize, y = DayLength_km,color = factor(Family))) +
  geom_point(na.rm = TRUE) +
  geom_smooth(method = "lm", se = FALSE, na.rm = TRUE) +
  facet_wrap(~ Family)
```
